---
title: "Differential CCA with dense paired data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Differential CCA with dense paired data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", dpi = 150,
                     fig.width = 10.67, fig.height = 5.67, fig.align = "center",
                     out.width = "100%")
library(dCCA)
```

## Generate paired dense observations

This reproducible example has two groups, each with 250 paired observations,
four X features, and four Y features. Every feature has a nonzero signal loading
and continuous noise. The latent variables for each paired observation are

\[
u \sim N(0,1), \qquad
v = r u + \sqrt{1-r^2}w, \qquad w \sim N(0,1).
\]

Their correlation r is 0.15 in the weak-coupling group and 0.85 in the
strong-coupling group. Observed feature vectors are generated as

\[
X_i = u_i a_0 + 0.55\varepsilon_{X,i}, \qquad
Y_i = v_i b_0 + 0.55\varepsilon_{Y,i},
\]

where the noise entries are independent standard normals, independent of the
latent variables. The shared unit-length loading vectors are normalized versions
of `(0.62, 0.49, -0.44, 0.42)` and `(0.55, -0.51, 0.48, 0.45)`.
Thus both groups have the same population marginal covariances, but different
cross-covariances: `Cov(X, Y) = r * a0 %*% t(b0)`. The latent generating
correlation is not the observed score correlation: feature noise attenuates it.
The generating loadings are never passed to dCCA.

## Fit common canonical weights

List elements are paired by position. `z = c(-1, 1)` contrasts the weak and
strong groups; dCCA standardizes these weights to approximately
`c(-0.707, 0.707)`. We leave `c1` and `c2` unset, giving dense weights without
L1 constraints. Eight paired starts reduce dependence on one initialization.
The default objective uses Fisher-transformed correlations; a second fit
explicitly uses raw correlations for comparison. Both fits estimate group
marginal covariances without assuming they are equal.

```{r generate-and-fit}
set.seed(475)
n <- 250
ax <- c(.62, .49, -.44, .42); ax <- ax / sqrt(sum(ax^2))
by <- c(.55, -.51, .48, .45); by <- by / sqrt(sum(by^2))
generate_group <- function(r) {
  latent_x <- rnorm(n)
  latent_y <- r * latent_x + sqrt(1 - r^2) * rnorm(n)
  list(X = outer(latent_x, ax) + .55 * matrix(rnorm(n * 4), n, 4),
       Y = outer(latent_y, by) + .55 * matrix(rnorm(n * 4), n, 4))
}
groups <- lapply(c(.15, .85), generate_group)
X <- lapply(groups, `[[`, "X"); Y <- lapply(groups, `[[`, "Y")
names(X) <- names(Y) <- c("Weak coupling", "Strong coupling")
a_starts <- c(list(rep(1, 4)), replicate(7, rnorm(4), simplify = FALSE))
b_starts <- c(list(rep(1, 4)), replicate(7, rnorm(4), simplify = FALSE))
fit <- dCCA(X = X, Y = Y, z = c(-1, 1),
  a_inits = a_starts, b_inits = b_starts, max_iter = 200)
raw <- dCCA(X = X, Y = Y, z = c(-1, 1),
  a_inits = a_starts, b_inits = b_starts, max_iter = 200,
  fisher.transform = FALSE)
```

## Inspect feature correlations and weights

The heatmaps summarize pairwise X-Y correlations before projection. The bar
plots show the fitted weights, which are used for both groups. Simultaneously
reversing both weight vectors leaves correlations unchanged.

```{r plot-helpers}
navy <- "#17324D"; teal <- "#168B83"; amber <- "#DD8A30"; muted <- "#657589"
ink <- "#25364A"; grid <- "#E6EBF0"

# Use consistent fonts and colors throughout the figures.
par_style <- function() {
  par(family = "sans", col = ink, col.axis = muted, col.lab = ink,
      col.main = navy, bg = "#FFFFFF", las = 1, bty = "n", tcl = -.2)
}
heatmap_panel <- function(i) {
  C <- cor(X[[i]], Y[[i]])
  colors <- colorRampPalette(c("#236893", "#F6F8FA", "#D67931"))(101)
  plot.new(); plot.window(xlim = c(.5, 4.5), ylim = c(.5, 4.5), asp = 1)
  for (j in 1:4) for (k in 1:4) {
    value <- C[j, k]
    rect(k - .48, 5 - j - .48, k + .48, 5 - j + .48,
         col = colors[round((value + 1) * 50) + 1], border = NA)
    text(k, 5 - j, sprintf("%.2f", value), cex = .95,
         col = if (abs(value) > .55) "white" else ink)
  }
  axis(1, at = 1:4, labels = paste0("Y", 1:4), tick = FALSE)
  axis(2, at = 1:4, labels = paste0("X", 4:1), tick = FALSE)
  title(main = names(X)[i], sub = "Pairwise X-Y feature correlations", cex.main = 1.1,
        col.sub = muted, cex.sub = .85)
}
```

```{r dense-input-and-weights, fig.height=8, echo=FALSE}
par_style()
par(mfrow = c(2, 2), mar = c(3.7, 4, 2.5, 1), oma = c(1.2, 0, 4.5, 0))
heatmap_panel(1); heatmap_panel(2)
for (view in c("X", "Y")) {
  w <- if (view == "X") fit$result$a else fit$result$b
  barplot(w, names.arg = paste0(view, 1:4), col = ifelse(w >= 0, teal, amber),
          border = NA, ylim = c(-1, 1), ylab = "Canonical weight",
          main = paste(view, "weights shared by both groups"))
  abline(h = 0, col = muted)
  mtext(if (view == "X") "u = X a" else "v = Y b", side = 1, line = 2.5, col = muted)
}
mtext("From dense input to a shared canonical direction", outer = TRUE,
      side = 3, line = 2.6, font = 2, cex = 1.45, col = navy)
mtext("Simulated data | 250 paired observations per group | 4 X + 4 Y features | no sparsity constraints",
      outer = TRUE, side = 3, line = .9, col = muted, cex = .95)
mtext("Weights are fitted using the default Fisher objective; simultaneous signs are arbitrary.",
      outer = TRUE, side = 1, line = .1, cex = .85, col = muted)
```

## Calculate and plot canonical scores

Within each group, center the observations and calculate `u = X a` and
`v = Y b`. These are the projected scores, not the latent variables used during
simulation. Score correlations computed directly agree with `fit$result$rho`.
The figure standardizes scores separately within each group for display;
this leaves their correlations unchanged.

```{r scores}
scores <- lapply(seq_along(X), function(i) {
  cbind(u = as.vector(scale(X[[i]], scale = FALSE) %*% fit$result$a),
        v = as.vector(scale(Y[[i]], scale = FALSE) %*% fit$result$b))
})
score_rho <- vapply(scores, function(s) cor(s[, 1], s[, 2]), numeric(1))
stopifnot(isTRUE(all.equal(score_rho, unname(fit$result$rho), tolerance = 1e-9)))
stopifnot(abs(sum(fit$parameters$z * atanh(score_rho)) - fit$result$obj_val) < 1e-9)
data.frame(group = names(X), correlation = score_rho)
```

```{r canonical-scores, fig.height=5.33, echo=FALSE}
par_style()
par(mfrow = c(1, 2), mar = c(4.7, 4.7, 3.2, 1), oma = c(2, 0, 4.2, 0))
for (i in 1:2) {
  s <- scale(scores[[i]])
  color <- c(amber, teal)[i]
  plot(s[, 1], s[, 2], pch = 16, cex = .8, col = adjustcolor(color, .45),
       xlim = c(-3.8, 3.8), ylim = c(-3.8, 3.8),
       xlab = "Standardized X score  (u = X a)",
       ylab = "Standardized Y score  (v = Y b)",
       main = sprintf("%s: rho = %.3f", names(X)[i], score_rho[i]))
  abline(h = 0, v = 0, col = grid)
  abline(a = 0, b = score_rho[i], col = color, lwd = 3)
}
mtext("The same weights expose different group correlations", outer = TRUE,
      side = 3, line = 2.4, cex = 1.45, font = 2, col = navy)
mtext("Each dot is one paired observation; lines show fitted score relationships.",
      outer = TRUE, side = 3, line = .8, cex = .95, col = muted)
mtext(sprintf("Observed correlation difference: rho(strong) - rho(weak) = %.3f", diff(score_rho)),
      outer = TRUE, side = 1, line = .6, cex = 1, col = navy)
```

The fitted score correlations are approximately 0.092 and 0.640, a difference
of 0.548. Finite sampling, feature noise, and optimization of a differential
objective account for differences from the latent correlations of 0.15 and 0.85.
Exact results can differ slightly across numerical-library versions.

## Verify the Fisher objective and convergence

For the selected component, dCCA maximizes

\[
J = \sum_g z_g\operatorname{atanh}(\rho_g).
\]

The objective is approximately `-0.707 * 0.092 + 0.707 * 0.757 = 0.471`.
The group contributions use transformed correlations; returned `rho` values
remain on the raw scale. Correlations are clipped just inside (-1, 1) before
transformation; this example is far from those boundaries.

```{r objective-calculation}
data.frame(
  group = names(X), z = fit$parameters$z,
  rho = score_rho, fisher_rho = atanh(score_rho),
  contribution = fit$parameters$z * atanh(score_rho)
)
fit$result$obj_val
fit$result$n_iter
fit$result$convergence
```

```{r objective-and-convergence, echo=FALSE}
par_style()
par(mfrow = c(1, 2), mar = c(4.7, 4.7, 3.3, 1), oma = c(2.4, 0, 4.2, 0))
trace <- fit$result$obj_val_record
plot(seq_along(trace) - 1, trace, type = "o", pch = 16, cex = .6,
     col = teal, lwd = 2.5, xlab = "Outer iteration (0 = initialization)",
     ylab = "Weighted Fisher objective", main = "Optimization of the selected start")
abline(h = fit$result$obj_val, col = muted, lty = 3)
legend("bottomright", legend = sprintf("%d iterations; convergence = %s",
       fit$result$n_iter, fit$result$convergence), bty = "n", cex = .85, text.col = muted)
z <- fit$parameters$z
fisher_rho <- atanh(score_rho)
contribution <- z * fisher_rho
barplot(contribution, names.arg = c("Weak group", "Strong group"),
        col = c(amber, teal), border = NA,
        ylim = range(c(0, contribution)) + c(-.12, .18),
        ylab = "Contribution to J = sum(z * atanh(rho))", main = "How the objective is calculated")
abline(h = 0, col = muted)
for (i in 1:2) {
  text(c(.7, 1.9)[i], contribution[i] + ifelse(contribution[i] >= 0, .075, -.075),
       sprintf("%.3f x %.3f = %.3f", z[i], fisher_rho[i], contribution[i]), cex = .9)
}
mtext("Fitting and calculating the differential objective", outer = TRUE,
      side = 3, line = 2.4, font = 2, cex = 1.45, col = navy)
mtext("Input z = (-1, +1) is standardized to (-0.707, +0.707); dense weights are estimated with CG.",
      outer = TRUE, side = 3, line = .8, cex = .95, col = muted)
mtext(sprintf("J = %.3f  |  Raw-objective fit on the same data: rho = (%.3f, %.3f)",
      fit$result$obj_val, raw$result$rho[1], raw$result$rho[2]), outer = TRUE,
      side = 1, line = .9, col = navy, cex = .95)
```

The displayed trace belongs to the selected start, not all eight fits. It
converges in about five outer iterations here. This is a numerical convergence
criterion, not proof of a global optimum. dCCA returns the best observed weights
and retains the full iteration history.

## Compare with raw-correlation fitting

```{r raw-comparison}
rbind(Fisher = fit$result$rho, Raw = raw$result$rho)
```

The raw-objective fit gives correlations of approximately 0.062 and 0.616.
The two objectives favor different weight directions; their numerical objective
values are on different scales and should not be ranked against one another.
Neither fit supplies a significance test or out-of-sample validation. For real
data, evaluate stability and generalization using independent observations.
